---
title: "Data Visualization Toolkit for PMA Panel Data"
description: |
  Heatmaps and alluvial plots make it easy to compare key family planning indicators over time and across multiple populations sampled by PMA panel surveys. 
author:
  - name: Matt Gunther
    affiliation: IPUMS PMA Senior Data Analyst
date: 2022-05-01
categories:
  - Panel Data
  - Data Visualization
  - Family Planning
  - Weights
  - heatmaps 
  - ggalluvial
  - ggplot2
  - srvyr
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
    toc_float: true
preview: images/long.png
---

```{r setup, echo=FALSE}
# knitr options 
knitr::opts_chunk$set(
  echo = TRUE, 
  eval = T,
  layout="l-page", 
  fig.width = 12,
  fig.height = 8,
  R.options = list(width = 100)
)

# tibble options 
options(tibble.print_min = 10)

# load utils 
source(here::here("r/utilities.r"))
set_postpath("2022-05-01-phase2-alluvial")
```

In our last post, we demonstrated [how to calculate](../2022-04-15-phase2-indicators/) key indicators for women's family planning status with new [panel surveys](../2022-03-01-phase2-discovery/index.html) from our partners at [PMA](https://www.pmadata.org/). When you download PMA data from [IPUMS PMA](https://pma.ipums.org/), you can include multiple samples together in the same data extract. Currently, Phase 1 and Phase 2 data are available for six samples, and we've devoted much of [this series](../../index.html#category:Panel_Data) to reproducing and synthesizing the major findings published in individual [PMA reports](https://www.pmadata.org/data/survey-results-summaries) for each sample: 

  * [Burkina Faso](https://www.pmadata.org/sites/default/files/data_product_results/Burkina%20National_Phase%202_Panel_Results%20Brief_English_Final.pdf)
  * [DRC - Kinshasa](https://www.pmadata.org/sites/default/files/data_product_results/DRC%20Kinshasa_%20Phase%202%20Panel%20Results%20Brief_English_Final.pdf)
  * [DRC - Kongo Central](https://www.pmadata.org/sites/default/files/data_product_results/DRC%20Kongo%20Central_%20Phase%202%20Panel%20Results%20Brief_English_Final.pdf)
  * [Kenya](https://www.pmadata.org/sites/default/files/data_product_results/Kenya%20National_Phase%202_Panel%20Results%20Brief_Final.pdf)
  * [Nigeria - Kano](https://www.pmadata.org/sites/default/files/data_product_results/Nigeria%20KANO_Phase%202_Panel_Results%20Brief_Final.pdf)
  * [Nigeria - Lagos](https://www.pmadata.org/sites/default/files/data_product_results/Nigeria%20LAGOS_Phase%202_Panel_Results%20Brief_Final.pdf) 

Now that we've learned how to compare population-level estimates for indicators in each population with [grouped bar charts](../2022-04-15-phase2-indicators/#visualization), we'd like to dig into some of the other data visualization tools that are commonly used for two-phase panel data. We'll quickly recap our approach to building bar charts, and then we'll explore color-coded crosstabs - or **heatmaps** - followed by **alluvial plots** built with [ggalluvial](https://corybrunson.github.io/ggalluvial), an extension to the [ggplot2](https://ggplot2.tidyverse.org) package for R. 

# Setup 

In addition to [ggalluvial](https://corybrunson.github.io/ggalluvial), we'll also showcase just a few of the helpful functions in [scales](https://scales.r-lib.org/), a package devoted to making axes, labels, and legends easier to read (in particular, we'll use [percent](https://scales.r-lib.org/reference/label_percent.html) to transform proportions into percentages labeled with the `%` symbol). You'll need those two packages, plus the three packages we've already used at length in this series: [tidyverse](https://tidyverse.tidyverse.org/), [ipumsr](https://tech.popdata.org/ipumsr/), and [srvyr](http://gdfe.co/srvyr).

```{r}
library(tidyverse)
library(ipumsr)
library(srvyr)
library(ggalluvial)
library(scales)
```

As a reminder, our last post featured a [wide format](../2022-03-15-phase2-formats/) data extract with "Female Respondents" only (other household members not participating in the panel study are excluded). If you're following along with this series, you can use the same extract again in this post; if not, you'll need to build a new extract that contains all of the variables we'll use here ([preselected](https://pma.ipums.org/pma-action/faq#ques28) variables are added automatically): 

  * `r varlink(RESULTFQ)` - Result of female questionnaire
  * `r varlink(PANELWEIGHT)` - Phase 2 female panel weight
  * `r varlink(RESIDENT)` - Household residence / membership
  * `r varlink(PREGNANT)` - Pregnancy status
  * `r varlink(GEOCD)` - Province, DRC
  * `r varlink(GEONG)` - State, Nigeria
  * `r varlink(CP)` - Contraceptive user
  * `r varlink(COUNTRY)` - PMA country (preselected)
  * `r varlink(EAID)` - Enumeration area (preselected)

Today's analysis will focus on three **recoded** variables we derived in our last post: 

  * `POP` - Population of interest 
  * `FPSTATUS_1` - Pregnant, using contraception, or using no contraception at Phase 1
  * `FPSTATUS_2` - Pregnant, using contraception, or using no contraception at Phase 2

Finally, our analysis dropped cases for women who only completed one of the two Female Questionnaires, were not members of the *de facto* population, or skipped critical questions regarding current use of family planning. If you'd like a refresher on the steps we took to load, filter, and recode data from our previous post, click the button below:

```{r, code_folding="Remind me about our data cleaning steps", results='hide'}
# import the data extract and metadata files
dat <- read_ipums_micro(
  ddi = "data/pma_00106.xml",
  data = "data/pma_00106.dat.gz"
) 

# inclusion criteria for analysis
dat <- dat %>% 
  filter(
    RESULTFQ_2 == 1,  # must have completed Phase 1 FQ
    RESIDENT_1 %in% c(11, 22) & # must be de facto population (both phases)
      RESIDENT_2 %in% c(11, 22),
    CP_1 < 90 & CP_2 < 90 # must answer "current FP use" question (both phases)
  ) 

# custom variables: `POP` and `FPSTATUS`
dat <- dat %>% 
  mutate(
    # Population of interest (country + region, where applicable)
    POP = case_when(
      !is.na(GEOCD) ~ paste("DRC -", as_factor(GEOCD)),
      !is.na(GEONG) ~ paste("Nigeria -", as_factor(GEONG)),
      TRUE ~ as_factor(COUNTRY) %>% as.character()
    ),
    # strata: includes placeholder values for DRC regions 
    STRATA_RECODE = if_else(
      is.na(GEOCD), 
      as.numeric(STRATA_1), 
      as.numeric(GEOCD)
    ),
    # Family planning use-status at Phase 1 
    FPSTATUS_1 = case_when(
      PREGNANT_1 == 1 ~ "Pregnant",
      CP_1 == 1 ~ "Using FP",
      CP_1 == 0 ~ "Not Using FP"
    ), 
    # Family planning use-status at Phase 2 
    FPSTATUS_2 = case_when(
      PREGNANT_2 == 1 ~ "Pregnant",
      CP_2 == 1 ~ "Using FP",
      CP_2 == 0 ~ "Not Using FP"
    ),
    # Create factors to control order of display in graphics output 
    across(
      c(FPSTATUS_1, FPSTATUS_2),
      ~.x %>% fct_relevel("Pregnant", "Not Using FP", "Using FP")
    )
  )

```

One more thing: our blog uses a particular font, layout, and color scheme that we've incorporated into a [custom theme](https://ggplot2.tidyverse.org/reference/index.html#themes) we call `theme_pma`. Feel free to use our theme, tweak it, or create your own. In case you're interested, we've included the code necessary to create our theme below (you'll need the [showtext](https://cran.rstudio.com/web/packages/showtext/vignettes/introduction.html) package to load a custom font). 

```{r, code_folding = "Remind me how we built `theme_pma`"}
# Custom font 
library(showtext)
sysfonts::font_add(
  family = "cabrito", 
  regular = "../../fonts/cabritosansnormregular-webfont.ttf"
)
showtext::showtext_auto()
update_geom_defaults("text", list(family = "cabrito", size = 4))

# Theme 
theme_pma <- theme_minimal() %+replace% 
  theme(
    text = element_text(family = "cabrito", size = 13),
    plot.title = element_text(size = 22, color = "#00263A",
    hjust = 0, margin = margin(b = 5)),
    plot.subtitle = element_text(hjust = 0, margin = margin(b = 10)),
    strip.background = element_blank(),
    strip.text.y = element_text(size = 16, angle = 0),
    panel.spacing = unit(1, "lines"),
    axis.title.y = element_text(angle = 0, margin = margin(r = 10))
  )
```

# Grouped Bar Charts

Now let's revisit the **grouped bar chart** we made to compare `FPSTATUS_1` and `FPSTATUS_2` for each population `POP`. We made this chart in basically two steps. 

First, we used [srvyr](http://gdfe.co/srvyr/index.html) to build a summary table that incorporates survey weights from `r varlink(PANELWEIGHT)` and generates a 95% confidence interval for each estimate. We used [EAID_1](https://pma.ipums.org/pma-action/variables/EAID#codes_section) to generate the cluster-robust standard errors underlying each confidence interval, and we stratified standard error estimation by `STRATA_RECODE`.

Notice that we [group_by](http://gdfe.co/srvyr/reference/group_by.html) `FPSTATUS_1` and `FPSTATUS_2` here. When we do this, [survey_mean](http://gdfe.co/srvyr/reference/survey_mean.html) estimates the proportion of outcomes represented by the variable that appears *last*, which is `FPSTATUS_2`. The proportions sum to `1.0` for each combination of `POP` and `FPSTATUS_1`: in other words, we obtain the proportion of `FPSTATUS_2` *on the condition* that women from a given `POP` held a particular status represented by `FPSTATUS_1`. For this reason, this is known as a **conditional distribution**. 

```{r}
status_tbl <- dat %>% 
  group_by(POP) %>% 
  summarise(
    .groups = "keep",
    cur_data() %>% 
      as_survey_design(weight = PANELWEIGHT, id = EAID_1, strata = STRATA_RECODE) %>%
      group_by(FPSTATUS_1, FPSTATUS_2) %>% 
      summarise(survey_mean(prop = TRUE, prop_method = "logit", vartype = "ci"))
  )

status_tbl
```

As a second step, we plotted each conditional distribution as a series of grouped bar charts arranged in [facets](https://ggplot2.tidyverse.org/reference/index.html#facetting) by `POP`. Because we wanted to recycle the same layout for several similar variables, we wrapped all of the necessary [ggplot2](https://ggplot2.tidyverse.org) tools together in a custom function called `pma_bars`.

```{r, code_folding = "Remind me how we built `pma_bars`"}
pma_bars <- function(
  title = NULL,     # an optional title 
  subtitle = NULL,  # an optional subtitle 
  xaxis = NULL,     # an optional label for the x-axis (displayed above)
  yaxis = NULL      # an optional label for the y-axis (displayed left)
){
  components <- list(
    theme_pma,
    labs(
      title = title,
      subtitle = subtitle,
      y = str_wrap(yaxis, 10),
      x = NULL,
      fill = NULL
    ),
    scale_x_continuous(
      position = 'bottom',
      sec.axis = sec_axis(trans = ~., name = xaxis, breaks = NULL),
      labels = scales::label_percent()
    ),
    scale_y_discrete(limits = rev),
    geom_bar(stat = "identity", fill = "#98579BB0"),
    geom_errorbar(
      aes(xmin = `_low`, xmax = `_upp`), 
      width = 0.2, 
      color = "#00263A"
    )
  )
}
```

```{r, fig.height=8}
status_tbl %>% 
  ggplot(aes(x = coef, y = FPSTATUS_2)) + 
  facet_grid(cols = vars(FPSTATUS_1), rows = vars(POP)) + 
  pma_bars(
    "CHANGE IN CONTRACEPTIVE USE OR NON-USE",
    "Percent women age 15-49 who changed contraceptive use status",
    xaxis = "Phase 1 Status",
    yaxis = "Phase 2 Status"
  )
```

# Heatmaps

We love this bar chart because it packs a lot of information into a single, reader-friendly graphic. However, we mentioned that it has some considerable drawbacks. Most importantly, we weren't able to include information from the **marginal distribution** in each phase. 

A marginal distribution for `FPSTATUS_1` would indicate the likelihood that a woman began the survey period pregnant, using family planning, or not using family planning. Likewise the marginal distribution for `FPSTATUS_2` estimates the likelihood that a woman would hold any such status at Phase 2, *independently of* her status at Phase 1. We call these distributions "marginal" because they're usually included in the row or column margins of a crosstab. 

Let's return to `status_tbl`, but this time we'll plot it as a crosstab with `color` and `alpha` (transparency) aesthetics. This type of crosstab is usually called a **heatmap**. First, we'll again wrap a few cosmetic layout options into a custom function we'll call `pma_heatmap`.

<aside>
For consistency, `pma_heatmap` uses the same color scheme shown in the first figure of each PMA report. 
</aside>

```{r, code_folding="Show me how `pma_heatmap` controls layout options"}
pma_heatmap <- function(
    title = NULL,     # an optional title 
    subtitle = NULL,  # an optional subtitle 
    xaxis = NULL,     # an optional label for the x-axis (displayed below)
    yaxis = NULL      # an optional label for the y-axis (displayed right)
){
  components <- list(
    theme_pma %+replace% theme(
      axis.text = element_text(size = 10),
      strip.text.x = element_text(size = 16,
                                  margin = margin(t = 10, b = 10)),
      axis.title.y.right = element_text(angle = 0, margin = margin(l = 10)),
      axis.title.x.bottom = element_text(margin = margin(t = 20)),
      panel.grid = element_blank(),
      legend.position = "none"
    ),
    labs(
      title = title,
      subtitle = subtitle,
      x = xaxis,
      y = str_wrap(yaxis, 10),
    ),
    scale_fill_manual(values = c(
      "Pregnant" = "#B4B3B3",
      "Not Using FP" = "#4E4F71",
      "Using FP" = "#EFD372"
    )),
    scale_color_manual(values = c("black", "white")),
    scale_y_discrete(position = "right", limits = rev)
  )
}
```

The plot, itself, is built with rectangles from [geom_tile](https://ggplot2.tidyverse.org/reference/geom_tile.html) and text labels from [geom_text](https://ggplot2.tidyverse.org/reference/geom_text.html). Both of these require a pair of x and y-coordinates, so we'll specify them globally in a [ggplot](https://ggplot2.tidyverse.org/reference/ggplot.html) function. 

Then, we tell [geom_tile](https://ggplot2.tidyverse.org/reference/geom_tile.html) to use one `fill` color for each type of response in `FPSTATUS_1`: this makes it easy for the reader to see that the totals in each tile sum to 100% in columns (not rows). The `alpha` aesthetic uses the value in `coef` to control the transparency of each color (by default, our minimum value `0` would be 100% transparent). 

The aesthetics in [geom_text](https://ggplot2.tidyverse.org/reference/geom_text.html) include `label` and `color`. Only `label` is really necessary: it tells the function to use the value in `coef` as a text label, except that we use [percent](https://scales.r-lib.org/) to stylize each number as a percentage rounded to the nearest integer. We're also including `color` to switch the font color from black to white for purple tiles where `coef` is higher than `0.5`: black text would be too hard to read here. 

Finally, [facet_wrap](https://ggplot2.tidyverse.org/reference/facet_wrap.html) plots each `POP` separately. We use `nrow = 3` to specify three rows, and we use `scales = "fixed"` to save a bit of space: the labels for `FPSTATUS_1` and `FPSTATUS_2` are printed only once in the row and column margins for the entire plot. 

```{r}
status_tbl %>% 
  ggplot(aes(x = FPSTATUS_1, y = FPSTATUS_2)) +
  geom_tile(aes(fill = FPSTATUS_1, alpha = coef)) + 
  geom_text(aes(
    label = scales::percent(coef, 1),
    color = coef > 0.5 & FPSTATUS_1 == "Not Using FP" # white vs black text
  )) + 
  facet_wrap(~POP, nrow = 3, scales = "fixed") + 
  pma_heatmap(
    "CHANGE IN CONTRACEPTIVE USE OR NON-USE",
    "Percent women age 15-49 who changed contraceptive use status",
    xaxis = "Phase 1 Status",
    yaxis = "Phase 2 Status"
  ) 
```

The nice thing about this heatmap layout is that we *can* easily include data from the marginal distribution of `FPSTATUS_1` and `FPSTATUS_2`. To do so, we'll first need to add them to `status_tbl`. Note: because our heatmap is not well-suited for comparing confidence intervals, we'll omit them in [survey_mean](http://gdfe.co/srvyr/reference/survey_mean.html) with the argument `vartype = NULL`.

```{r}
col_margins <- dat %>% 
  group_by(POP) %>% 
  summarise(
    .groups = "keep",
    cur_data() %>% 
      as_survey_design(weight = PANELWEIGHT, id = EAID_1, strata = STRATA_RECODE) %>%
      group_by(FPSTATUS_1) %>% 
      summarise(cols = survey_mean(prop = TRUE, prop_method = "logit", vartype = NULL))
  )
  
row_margins <- dat %>% 
  group_by(POP) %>% 
  summarise(
    .groups = "keep",
    cur_data() %>% 
      as_survey_design(weight = PANELWEIGHT, id = EAID_1, strata = STRATA_RECODE) %>%
      group_by(FPSTATUS_2) %>% 
      summarise(rows = survey_mean(prop = TRUE, prop_method = "logit", vartype = NULL))
  )

status_tbl <- status_tbl %>% right_join(col_margins) %>% right_join(row_margins)

status_tbl
```

To keep things simple, we've named the marginal distribution for `FPSTATUS_1` "cols", and the marginal distribution for `FPSTATUS_2` "rows". Ultimately, we think it's clearest to [paste](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/paste) these values (as percentages) together with the original labels from `FPSTATUS_1` and `FPSTATUS_2` (the symbol `\n` represents a line break). We'll also transform each character string into a [factor](https://forcats.tidyverse.org/reference/as_factor.html), which ensures that the values will be plotted in the same order that they appear in our table. 

Finally, we switch from `scales = "fixed"` to `scales = "free"`. This time, we'll want to print the row and column margins for each `POP` separately. 

```{r, fig.height=9}
status_tbl %>% 
  ggplot(aes(
    x = paste0(scales::percent(cols, 1), "\n", FPSTATUS_1) %>% as_factor, 
    y = paste0(scales::percent(rows, 1), "\n", FPSTATUS_2) %>% as_factor
  )) +
  geom_tile(aes(fill = FPSTATUS_1, alpha = coef)) + 
  geom_text(aes(
    label = scales::percent(coef, 1),
    color = coef > 0.5 & FPSTATUS_1 == "Not Using FP"
  )) + 
  facet_wrap(~POP, nrow = 3, scales = "free") + 
  pma_heatmap(
    "CHANGE IN CONTRACEPTIVE USE OR NON-USE",
    "Percent women age 15-49 who changed contraceptive use status",
    xaxis = "Phase 1 Status",
    yaxis = "Phase 2 Status"
  )
```

The information contained in our heatmap is similar to what we saw in our bar chart, except for two things:

  1. There are no error bars on our heatmap. If we wanted to include information about the confidence interval for each estimation, we would have to include [text symbols](https://search.r-project.org/CRAN/refmans/gtools/html/stars.pval.html).
  2. While both plots show information about the conditional distribution of `FPSTATUS_2` given a starting point in `FPSTATUS_1`, only the heatmap includes the marginal distribution of each variable in its row and column margins. 
  
The marginal distribution may provide crucial information about the conditional distribution that we would otherwise miss. Consider Burkina Faso, where both users and non-users of family planning at Phase 1 were generally most likely to maintain their status at Phase 2. The marginal distribution adds additional information: non-users comprise a larger share of the overall population at Phase 1.    

In certain contexts, you may want to combine information from the Phase 1 marginal distribution together with the conditional distribution of outcomes at Phase 2. To continue with our example from Burkina Faso, you might report that - because
non-users represent about 62% of the population, only about 11% of the population adopted family planning at Phase 2 following non-use at Phase 1. That is: 18% of 62% is 11%. 

In contrast with the conditional distribution, this type of distribution describes the share of the population that experiences some combination of Phase 1 and Phase 2 outcomes *without* assuming a particular starting point at Phase 1. It's known as a **joint distribution** because it gives the probability that two events will happen together (in sequence). Let's return to our summary table, `status_tbl`: 

```{r}
status_tbl
```

To find the estimated joint distribution for each combination of `FPSTATUS_1` and `FPSTATUS_2`, you could simply multiply each value in `cols` by the value in `coef`: 

```{r, R.options = list(width = 100)}
status_tbl %>% mutate(joint = cols * coef)
```

In practice, you'll usually want to let [srvyr](http://gdfe.co/srvyr) calculate a confidence interval for each joint probability. To do so, we'll add an [interact](http://gdfe.co/srvyr/reference/interact.html) function listing the variables in [group_by](http://gdfe.co/srvyr/reference/group_by.html) that we want to model jointly. 

```{r, R.options = list(width = 100)}
joint_tbl <- dat %>% 
  group_by(POP) %>% 
  summarise(
    .groups = "keep",
    cur_data() %>% 
      as_survey_design(weight = PANELWEIGHT, id = EAID_1, strata = STRATA_RECODE) %>%
      group_by(interact(FPSTATUS_1, FPSTATUS_2)) %>% 
      summarise(joint = survey_mean(prop = TRUE, prop_method = "logit", vartype = "ci"))
  )
  
joint_tbl
```

Now, the values in `joint` sum to `1.0` for each `POP`. Returning to our heatmap, we'll want to use the same color for every column, indicating that the percentages sum for 100% for each population. 

```{r}
joint_tbl %>% 
  ggplot(aes(x = FPSTATUS_1, y = FPSTATUS_2)) +
  geom_tile(aes(alpha = joint), fill = "#98579B") + 
  geom_text(aes(
    label = scales::percent(joint, 1),
    color = joint > 0.5 & FPSTATUS_1 == "Not Using FP"
  )) + 
  facet_wrap(~POP, nrow = 3, scales = "fixed") + 
  pma_heatmap(
    "CHANGE IN CONTRACEPTIVE USE OR NON-USE",
    "Percent women age 15-49 who changed contraceptive use status",
    xaxis = "Phase 1 Status",
    yaxis = "Phase 2 Status"
  ) 
```

Information provided by the joint distribution nuances our story a bit further. To continue with our examination of Burkina Faso: we knew that family planning users and non-users at Phase 1 were each most likely to maintain, rather than switch their status at Phase 2. However, it's now clear that *continuous non-users* (non-users at both Phase 1 and Phase 2) represent a near-majority of the population.

# Alluvial plots 

**Alluvial plots** are an especially popular way to visualize longitudinal data, in part, because they combine information from each of the three distributions we've discussed. They also make it possible to show data from more than two variables (we'll use them again when Phase 3 data become available). You'll find alluvial plots on the first two pages of the PMA report for each sample. 

In an alluvial plot, the marginal distribution of responses for each variable are usually plotted in vertical stacks. The [ggalluvial](https://corybrunson.github.io/ggalluvial/index.html) package authors refer to these stacks as "strata", and they may be layered onto a [ggplot](https://ggplot2.tidyverse.org/reference/ggplot.html) with [geom_stratum](https://corybrunson.github.io/ggalluvial/reference/geom_stratum.html). In our case, the strata will show the marginal distribution of women in `FPSTATUS_1` and `FPSTATUS_2`. 

The **joint distribution** for any pair of variables is plotted in horizontal splines called "alluvia", which bridge the space between any given pair of strata. Alluvia are plotted with [geom_flow](https://corybrunson.github.io/ggalluvial/reference/geom_flow.html). 

Finally, we'll use color to map each alluvium with an originating stratum from `FPSTATUS_1`. This will help the reader visualize the conditional distribution of `FPSTATUS_2` responses given a starting point in `FPSTATUS_1`. 

To begin, let's revisit `joint_tbl`, which only contains the joint distribution for `FPSTATUS_1` and `FPSTATUS_2`. In fact, [ggalluvial](https://corybrunson.github.io/ggalluvial/index.html) will calculate the marginal distribution for both variables automatically if we reshape `joint_tbl` with [pivot_longer](https://tidyr.tidyverse.org/reference/pivot_longer.html) like so: 

```{r, R.options = list(width = 100)}
joint_tbl <- joint_tbl %>% 
  rowid_to_column("alluvium") %>% 
  pivot_longer(
    c(FPSTATUS_1, FPSTATUS_2), 
    names_to = "x", 
    values_to = "stratum"
  ) %>% 
  mutate(x = ifelse(x == "FPSTATUS_1", "Phase 1", "Phase 2")) %>% 
  arrange(x, alluvium)

joint_tbl
```

Here, we create the column `alluvium` to hold the original row number for each of the 56 combinations of `POP`, `FPSTATUS_1`, and `FPSTATUS_2`. When we [pivot_longer](https://tidyr.tidyverse.org/reference/pivot_longer.html), we repeat the value in `joint` once for each end of the same `alluvium`. The values in `stratum` describe the strata to to which each alluvium is attached, and `x` indicates whether the stratum is located in the Phase 1 or Phase 2 stack. 

As with our heatmap, we'll want to define some custom fonts, color, and layout options adapted from `theme_pma`. We'll bundle these together in a function called `pma_alluvial` - feel free to use, adjust, or omit this function for your own purposes. 

<aside>
For consistency, `pma_alluvial` uses the same color scheme shown in the first alluvial plot in each PMA report.
</aside>

```{r, code_folding="Show me how `pma_alluvial` controls layout options"}
pma_alluvial <- function(
    title = NULL,     # an optional title 
    subtitle = NULL,  # an optional subtitle 
    xaxis = NULL,     # an optional label for the x-axis (displayed below)
    yaxis = NULL      # an optional label for the y-axis (displayed left)
){
  components <- list(
    theme_pma %+replace% theme(
      plot.title = element_text(size = 22, color = "#541E5A", 
                                hjust = 0.5, mar = margin(b = 5)),
      plot.subtitle = element_text(hjust = 0.5, margin = margin(b = 20)),
      axis.text.x = element_text(color = "#541E5A", 
                                 margin = margin(t = 5, b = 10)),
      strip.text.x = element_text(size = 13, margin = margin(b = 5)),
      plot.margin = margin(0, 100, 0, 100),
      legend.position = "bottom",
      legend.title = element_blank(),
      legend.spacing.x = unit(10, "pt"),
      panel.grid = element_blank(),
      axis.text.y = element_blank()
    ),
    labs(
      title = title,
      subtitle = subtitle,
      x = xaxis,
      y = str_wrap(yaxis, 10),
    ),
    scale_fill_manual(values = c(
      "Pregnant" = "#B4B3B3", 
      "Not Using FP" = "#4E4F71", 
      "Using FP" = "#EFD372"
    )),
    scale_y_continuous(expand = c(0, 0))
  )
}
```

We'll start by mapping common aesthetics in a [ggplot](https://ggplot2.tidyverse.org/reference/ggplot.html) function. We'll map the values in `x` onto our x-axis, and we'll map the values in `joint` onto the y-axis. The remaining aesthetics are specific to the functions from [ggalluvial](https://corybrunson.github.io/ggalluvial/index.html): we'll use `stratum` to build vertical strata and to define colors mapped with "fill". We also use the identifying numbers in `alluvium` to organize responses into alluvia.

The remaining functions are straightforward, since they mainly use information passed from [ggplot](https://ggplot2.tidyverse.org/reference/ggplot.html). We make only one small modification to [geom_stratum](https://corybrunson.github.io/ggalluvial/reference/geom_stratum.html): setting `size = 0` removes border lines that appear around each stratum, by default. 

```{r, layout = "l-screen", fig.width=18, fig.height=8, message=FALSE}
status_alluvial <- joint_tbl %>% 
  ggplot(aes(
    x = x, 
    y = joint,
    fill = stratum,
    stratum = stratum,
     alluvium = alluvium
  )) + 
  geom_flow() + 
  geom_stratum(size = 0) + 
  facet_wrap(~POP, scales = "free_x", nrow = 1) + 
  pma_alluvial(
    "CHANGE IN CONTRACEPTIVE USE OR NON-USE",
    "Percent women age 15-49 who changed contraceptive use status",
  ) 

status_alluvial
```

Of course, you should always include either y-axis gridlines or text labels for the probabilities shown on a plot like this one. We find it clearer to include the latter, which we'll build with [geom_text](https://ggplot2.tidyverse.org/reference/geom_text.html). 

These labels are a bit tricky, but the basic idea is that you use `stat = "stratum"` to label strata, and `stat = "flow"` to label alluvia. Then, you use [after_stat](https://ggplot2.tidyverse.org/reference/aes_eval.html) to build labels from statistics that [ggalluvial](https://corybrunson.github.io/ggalluvial/index.html) uses to construct the plot - check out [this list](https://corybrunson.github.io/ggalluvial/reference/stat_stratum.html#computed-variables) of available statistics for details. We'll use the `prop` statistic to obtain *both* the marginal and joint probabilities for each outcome (we'll leave the conditional probabilities unlabeled, but you could adjust this code to include them here).

<aside>
Values may not add to 100% due to rounding (values rounded to 0% are not labelled).
</aside>

```{r, layout = "l-screen", fig.width=18, fig.height=8}
status_alluvial +
  geom_text(
    stat = "stratum",   # label strata
    aes(label = ifelse( 
      x == 1, # labels the strata for Phase 1, otherwise blank ""           
      scales::percent(after_stat(prop), 1), 
      ""
    )),
    nudge_x = -0.2,  # nudge a bit to the left 
    hjust = "right", # right-justify
  ) + 
  geom_text(
    stat = "stratum", # label strata
    aes(label = ifelse(
      x == 2, # labels the strata for Phase 2, otherwise blank ""    
      scales::percent(after_stat(prop), 1), 
      ""
    )),
    nudge_x = 0.2,  # nudge a bit to the right 
    hjust = "left", # left-justify
  ) +
  geom_text(
    stat = "flow",   # label alluvia 
    aes(label = ifelse( 
      after_stat(flow) == "to" &  # only label the destination (right-side)
        after_stat(prop) >= 0.01, # hide if 0%
      scales::percent(after_stat(prop), 1), 
      "" 
    )),
    nudge_x = -0.2,  # nudge a bit to the left
    hjust = "right", # right-justify
    size = 3         # use a slightly smaller font 
  ) 
```

Now, it's easy to identify the proportion of women at each phase *and* the proportion who switched or maintained their status between phases. If possible, we recommend aligning alluvial plots for every sample in a single row as shown: this allows the readers to visually compare the relative size of strata and alluvia across samples. 

# Next Steps 

So far in [this series](../../index.html#category:Panel_Data), we've covered topics related to: 

  * [Data availability](../2022-03-01-phase2-discovery/)
  * [Instructions of obtaining data](../2022-03-15-phase2-formats/) 
  * [Sample design](../2022-04-01-phase2-members/) 
  * [Key family planning indicators](../2022-04-15-phase2-indicators/) 
  * [Data Visualization](../2022-05-01-phase2-alluvial/) 
  
In two weeks, we'll be wrapping up this introduction to PMA Panel Data with an update on a topic we first covered last year: the [contraceptive calendar](https://tech.popdata.org/pma-data-hub/posts/2021-05-15-paa-2021/) section of the Female Questionnaire. As we'll see, the calendar adds a different temporal dimension to the panel study: it represents the contraceptive use, non-use, and pregnancy status of women recalled on a *monthly basis* for several months prior the interview. We'll show how to parse these data from string-format, and how to merge responses obtained at Phase 1 and Phase 2 for each woman. 