---
title: "Visualizing PMA Data with StoryMaps"
description: |
  Five outstanding undergraduate research projects integrate dynamic data visualization with spatial analysis and narrative. 
author:
  - name: Matt Gunther
    affiliation: IPUMS PMA Senior Data Analyst
date: 05-03-2021
categories:
  - PMA Publications
  - StoryMaps
  - Undergrads
  - Teaching
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
    toc_float: true
preview: images/featured.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

When great new research gets published with PMA data, we like to share it with you in a [PMA Publications](../../index.html#category:PMA_Publications) post. Sometimes we'll take a deep dive into the R code you can use to [reproduce great analysis](../2021-04-01-et-internal-migration/index.html), but today we'll take a look at a different tool that undergraduate students are using to learn about IPUMS PMA and [other IPUMS Global Health data](https://globalhealth.ipums.org/) at the University of Minnesota.

<aside>
Have a recent publication using PMA data that you’d like to feature in a [PMA Publications](../../index.html#category:PMA_Publications) post? Please [let us know](https://github.com/ipums/pma-data-hub/issues/new)!
</aside>

This semester, students in the [Global Health Survey Analysis](https://umn.maps.arcgis.com/home/group.html?id=e977beb52a9e4169a0ba5584336cd934&view=list#content) course used an amazing tool called [StoryMaps](https://storymaps.umn.edu/) to develop interactive narratives exploring different topics related to family planning. StoryMaps have been used in both the undergraduate and graduate curriculum throughout the College of Liberal Arts and beyond - we encourage you to check out the full gallery of student projects [here](https://storymaps.umn.edu/pages/gallery)!

*Thanks to course professors [Elizabeth Boyle](https://cla.umn.edu/about/directory/profile/boyle014) and [Kathryn Grace](https://www.kathryn-q-grace.com/) for sharing this great work!*

# Attainment of Sex Preference in India

Author: Lara Rae Erdmann

```{r, layout="l-page"}
knitr::include_url(
  "https://storymaps.arcgis.com/stories/6fcdc8dcd95b4ca99b891427b9f38a8b",
  height = "650px"
)
```

# Intimate Partner Violence and Body Weight

Author: Jaclyn Willems

```{r, layout="l-page"}
knitr::include_url(
  "https://storymaps.arcgis.com/stories/2a714b3729db49e9841f016e45ccd0bd", 
  height = "650px"
)
```

# Higher Probability that Women Report IPV

Author: Peyton Retterath

```{r, layout="l-page"}
knitr::include_url(
  "https://storymaps.arcgis.com/stories/0f265dca10c9450c8586fb0219a92321",
  height = "650px"
)
```

# Empowered Women Raise Healthy Children

Author: Kassandra Fate

```{r, layout="l-page"}
knitr::include_url(
  "https://storymaps.arcgis.com/stories/31c34bc1e828494aaca9dea7c42bceb9",
  height = "650px"
)
```

# Wealth and Healthcare can Save Pregnant Women

Author: Hana al'Absi

```{r, layout="l-page"}
knitr::include_url(
  "https://storymaps.arcgis.com/stories/84d2ff5d2cd24d1da380e95913f4bd08",
  height = "650px"
)
```
